---
title: "HW2 Prince George's County Crime"
---

In this assignment, we will be working with a dataset reporting the number of crimes by month in Prince George's County, Maryland, for every month in 2024. 

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv. You will create a dataframe called `pg_crime` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
pg_crime <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, are any of the categories more important than others? Why or why not? Do any of them make the job of summarizing the data more difficult? Why or why not?

After looking at the data and seeing the range of crime, more serious crimes, like homicide and assault are more worthy of reporting. Violent crime carries a lot more public weight and engagement, and it's important to know about. There are a few things that complicate summarizing the data cleanly, and the major one is the overlap. One category is called "robbery commercial" and another one is called "robbery other." I think before getting a clear picture, the data needs to be synthesized a bit more for organization. 

### Task 4: Calculating Averages

Add up the total number of incidents per month for each category using `mutate()`, calling the new column `total_incidents`. Then calculate the average number of incidents per month for each category of crime using the total number of months, calling the new column `average_incidents`. Assign the result to a new dataframe called `pg_crime_averages`.

```{r}
pg_crime_averages <- mutate(pg_crime,
                             total_incidents = Jan + Feb + Mar + Apr + May + Jun + Jul + Aug + Sep + Oct + Nov + Dec,
                             average_incidents = total_incidents / 12)
```

### Task 5: Describe the Averages

Examine the new dataframe, `pg_crime_averages`, by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the average number of incidents per month for each category. Which categories have the highest average number of incidents per month? Which have the lowest? What does this tell you about the pattern of crime in Prince George's County? Are there categories that you have questions about? Why or why not?

The categories with the highest number of incidents on average is accidents, accidents with impound, theft from auto and stolen auto. I am a bit confused why these can't be synthesized into two categories: auto theft and accidents, but it makes sense to me that these are the highest. Based on my notifications from UMPD, I often see auto-related crimes. I also know people who have had their cars stolen, so it must be common. The ones with the lowest is b and e and other and homicide. I am certainly relieved that homicide is low, but I wonder if it includes suicide. I am curious what NA means in this context. 


### Task 6: Testing a Theory

One theory about crime is that it is more likely to occur in the summer months. Test this theory by creating three new columns using `mutate()` that calculate the percentage change compared to the average for each category of crime for the months of June, July, and August. Call the new columns `june_change`, `july_change`, and `august_change`. Assign the result to a new dataframe called `summer_crime_averages`.

```{r}
summer_crime_averages <- mutate(pg_crime_averages,
                                 june_change = (Jun - average_incidents) / average_incidents * 100,
                                 july_change = (Jul - average_incidents) / average_incidents * 100,
                                 august_change = (Aug - average_incidents) / average_incidents * 100)
```

### Task 7: Describe the Summer Crime Averages

Look at the `summer_crime_averages` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change in the average number of incidents per month for each category of crime for the months of June, July, and August. Do you see any patterns? What does this tell you about the theory that crime is more likely to occur in the summer months?

It appears homicide rises in the summer, which is pretty concerning. Assualt shooting and sex offense also rose for the summer months, indicating a rise in violent crimes in the summer. I wonder if it's in just the summer, or any time when the student body is not around, like winter break. The one with the most uptick is robbery vehicle, which makes the most sense to me. 

### Task 8: What's the Story?

Write up a lede and nutgraf of a story focused on the crime data in Prince George's County. It could be from any of the dataframes you've created in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Before University of Maryland students bathe in the Punta Cana sun and party until sunrise in Miami this summer, they should be aware of the crime that spikes while they're away.

While the student population declines over the summer, crime doesn't follow the same pattern. An analysis of county crime data from 2024 shows that while overall incident totals didn't rise dramatically in the summer, violent crime did; homicides, shootings and robberies all climbed well above their yearly average in June, July and August, suggesting Prince George's County's summer crime problem is more complex than a simple drop in students.

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
