---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2026.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2026.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
url <- "https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2026.csv"

wh_salaries <- read_csv(url)

```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

It contains 411 White House staffers with a wide salary range. The range for entry level positions to senior positions is from $59,661 to $197,200. One issue I see is a senior position with a salary of 0. This will skew median calculations. 


### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |> 
  summarize(mean_salary = mean(SALARY, na.rm = TRUE), 
            median_salary = median(SALARY, na.rm = TRUE))
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |> 
  summarize(mean_salary = mean(SALARY, na.rm = TRUE), 
            median_salary = median(SALARY, na.rm = TRUE),
            min_salary = min(SALARY, na.rm = TRUE),
            max_salary = max(SALARY, na.rm = TRUE))
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

The mean salary is about $112,703 and the median is $109,243. Since they are close together, it suggests the data is not widely skewed. The 0s are the clear outliers. 

### Task 7: Story Idea
Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?
Several high profile White House advisors, including Steven Witkoff, are serving without pay. I'd be curious to learn more why they aren't drawing a salary and the ethics behind free labor. My next step would be cross referencing the $0-salary staffers and detailees against public financial disclosure filings.
