---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
library(janitor)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities |> 
  mutate(violent_rate_2019 = violent2019/pop2019 * 10000, 
         property_rate_2019 = property2019/pop2019 * 10000,
         violent_rate_2020 = violent2020/pop2020 * 10000,
         property_rate_2020 = property2020/pop2020 * 10000)
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?


From what I see, Ocean City, Oakland, Elkton, Hyattsville and Cambridge are highest for propery crime rate in 2019. This is interesting because they are all midsize cities, and you'd expect it to be bigger ones, like Baltimore. Showing the importance of rate-per-population. There are aksi a lot of tows with 0 for their rates, showing an uneven crime reportinf across Maryland. 

### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities_rates |> 
  mutate(violent_pct_change = (violent_rate_2020 - violent_rate_2019)/violent_rate_2019 * 100,
         property_pct_change = (property_rate_2020 - property_rate_2019)/property_rate_2019 * 100)
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

The outliers stood out to me immediately. Edmonston's violent crime rate rose nearly 900% and Crisfield's property rate rose over 300%. But, these these are smaller towns, where a few additional incidents can cause an enormous percentage change without reflecting any real shift in public safety.Overall, Property crime rates declined two-thirds of Maryland cities between 2019 and 2020, with a median drop of about 15%, which makes sense with the pandemic lockdowns reducing opportunities for crimes like burglary and theft. Violent crime rates stayed roughly flat overall, with a median change close to 0%, indicating the pandemic didn't affect all crime categories the same way.

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

The pandemic caused property crime rates to fall in nearly two-thirds of Maryland's cities and towns between 2019 and 2020.

The decline shows that while the pandemic caused devastation in other ways, it reshaped opportunity for property crimes like burglary and theft far more than it did violent crime, which held roughly steady statewide even as daily life ground to a halt. The pattern held broadly across cities big and small, from Baltimore to small towns in Western Maryland, pointing to a statewide shift rather than a handful of isolated success stories. This means crime isn't just about policing or policy, but also about the everyday routines, like commuting, shopping and socializing, that create chances for crime to happen. When those routines disappeared in 2020, a meaningful share of the property crimes that depend on them did too.

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

#### Task 10: take the violent crime rates for 2020 and do three things

```{r}
md_cities_percentiles <- md_cities_rates |> 
  mutate(percentile = ntile(violent_rate_2020, 100))
```

```{r}
md_cities_percentiles |> filter(percentile > 90)
```

```{r}
md_cities_rates |> summarize(violent_iqr_2020 = IQR(violent_rate_2020))
```
The 90th percentile results support my earlier story. Ocean City and Baltimore top the list, along with several small towns like Cambridge and Princess Anne, which helps the idea that small population and tourist traffic can raise crime rates. The IQR of 42 shows most Maryland cities are close together, making these 14 cities are clear outliers. This adds a useful detail to the story: violent crime in 2020 was concentrated in a small number of places, while the broader statewide trend was mainly a decline in property crime.

